---
title: "Mini-Project #01"
editor: visual
---
Author: Thanh Dao\
Updated: October 3rd, 2024

## Introduction

There is a great variety of transportation systems and modes in America, spanning across land and water. Americans utilize all of them on a daily basis, commuting to wherever they need to go. In this report, the fiscal characteristics of major US public transit systems will be analyzed.

## Obtaining Data

This report runs on R; thus, the necessary packages must be installed prior to any data collection or analysis.

```{r echo=TRUE, message=FALSE}
# Install the necessary R packages
if(!require("tidyverse")) install.packages("tidyverse")
library(lubridate)  # added for the my() function
if(!require("DT")) install.packages("DT")
library(DT)
```

Now that the necessary packages have been installed, data collection from the [National Transit Database](https://www.transit.dot.gov/ntd) can proceed.

```{r echo=TRUE, message=FALSE}
# Download the "Fare Revenue" Data
if(!file.exists("2022_fare_revenue.xlsx")){
  download.file("http://www.transit.dot.gov/sites/fta.dot.gov/files/2024-04/2022%20Fare%20Revenue.xlsx", 
                destfile="2022_fare_revenue.xlsx", 
                quiet=FALSE, 
                method="wget")
}

# Select the columns within the scope of research for this report
FARES <- readxl::read_xlsx("2022_fare_revenue.xlsx") |>
  select(-`State/Parent NTD ID`, 
         -`Reporter Type`,
         -`Reporting Module`,
         -`TOS`,
         -`Passenger Paid Fares`,
         -`Organization Paid Fares`) |>
  filter(`Expense Type` == "Funds Earned During Period") |>
  select(-`Expense Type`)

# Download the "Expenses" Data
if(!file.exists("2022_expenses.csv")){
  download.file("https://data.transportation.gov/api/views/dkxx-zjd6/rows.csv?date=20231102&accessType=DOWNLOAD&bom=true&format=true", 
                destfile="2022_expenses.csv", 
                quiet=FALSE, 
                method="wget")
}

# Select the columns within the scope of research for this report
EXPENSES <- readr::read_csv("2022_expenses.csv") |>
  select(`NTD ID`, 
         `Agency`,
         `Total`, 
         `Mode`) |>
  mutate(`NTD ID` = as.integer(`NTD ID`)) |>
  rename(Expenses = Total) |>
  group_by(`NTD ID`, `Mode`) |>
  summarize(Expenses = sum(Expenses)) |>
  ungroup()

# Combine the selected "Fare Revenue" and "Expenses" Data
FINANCIALS <- inner_join(FARES, EXPENSES, join_by(`NTD ID`, `Mode`)) 

# Download the "Monthly Transit Numbers" Data
if(!file.exists("ridership.xlsx")){
  download.file("https://www.transit.dot.gov/sites/fta.dot.gov/files/2024-09/July%202024%20Complete%20Monthly%20Ridership%20%28with%20adjustments%20and%20estimates%29_240903.xlsx", 
                destfile="ridership.xlsx", 
                quiet=FALSE, 
                method="wget")
}

# Select the columns within the scope of research for this report
TRIPS <- readxl::read_xlsx("ridership.xlsx", sheet="UPT") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="UPT") |>
  drop_na() |>
  mutate(month=my(month)) 
MILES <- readxl::read_xlsx("ridership.xlsx", sheet="VRM") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="VRM") |>
  drop_na() |>
  group_by(`NTD ID`, `Agency`, `UZA Name`, 
           `Mode`, `3 Mode`, month) |>
  summarize(VRM = sum(VRM)) |>
  ungroup() |>
  mutate(month=my(month))

# Combine the selected "Monthly Transit Numbers" Data
USAGE <- inner_join(TRIPS, MILES) |>
  mutate(`NTD ID` = as.integer(`NTD ID`))
```

## Initial Data Table

Using the data just obtained from the National Transit Database, the following table can be created:

```{r echo=FALSE, message=FALSE}
sample_n(USAGE, 1000) |> 
    mutate(month=as.character(month)) |> 
    DT::datatable() 
```

\
Unfortunately, this data still some flaws that need to be fixed.

## Cleaning the Data

To the average person, "UZA Name" doesn't provide proper context or meaning; thus, it needs to be changed to something people will understand, like "Metro Area." The same can be said about "UPT" and "VRM"; these also need to be changed, as follows: 

```{r echo=TRUE, message=FALSE}
# Task 1: Create Syntactic Names
USAGE <- rename(USAGE, "metro_area" = "UZA Name",
                "unlinked_passenger_trips" = "UPT",
                "vehicle_revenue_miles" = "VRM",
                "NTD_ID" = "NTD ID")
FINANCIALS <- rename(FINANCIALS, "NTD_ID" = "NTD ID")
```

Now that some column headers have been renamed, focus can be made on the "Mode" column. All of these modes of transport are acronyms that would need to be looked up to be understood. For user ease, they'll be changed as follows: 

```{r echo=TRUE, message=FALSE}
# Task 2: Recoding the Mode Column
USAGE <- USAGE |>                    # Interpret the Mode column in "USAGE"
  mutate(Mode = case_when(
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "CC" ~ "Cable Car",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferryboat",
    Mode == "HR" ~ "Heavy Rail",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "LR" ~ "Light Rail",
    Mode == "MB" ~ "Bus",
    Mode == "MG" ~ "Monorail and Automated Guideway",
    Mode == "PB" ~ "Publico",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "SR" ~ "Streetcar Rail",
    Mode == "TB" ~ "Trolleybus",
    Mode == "TR" ~ "Aerial Tramways",
    Mode == "VP" ~ "Vanpool",
    Mode == "YR" ~ "Hybrid Rail",
    TRUE ~ "Unknown"))

FINANCIALS <- FINANCIALS |>          # Interpret the Mode column in "Financials"
  mutate(Mode = case_when(
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "CC" ~ "Cable Car",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferryboat",
    Mode == "HR" ~ "Heavy Rail",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "LR" ~ "Light Rail",
    Mode == "MB" ~ "Bus",
    Mode == "MG" ~ "Monorail and Automated Guideway",
    Mode == "PB" ~ "Publico",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "SR" ~ "Streetcar Rail",
    Mode == "TB" ~ "Trolleybus",
    Mode == "TR" ~ "Aerial Tramways",
    Mode == "VP" ~ "Vanpool",
    Mode == "YR" ~ "Hybrid Rail",
    TRUE ~ "Unknown"))
```

Unnecessary columns will be removed as well.

```{r echo=TRUE, message=FALSE}
USAGE <- USAGE |>           # Remove "3 Mode"
  select(-c("3 Mode"))
```

Now that the data is cleaned up, a new table can be established, as follows:\

```{r echo=FALSE, message=FALSE}
sample_n(USAGE, 1000) |>    # Updated Summary Table
  mutate(month=as.character(month)) |> 
  DT::datatable()
```

## Main Analysis

Using this table, the following analyses can be done.

The **transit agency with the most total VRM** in this data set can be determined with the following code: 

```{r echo=TRUE, message=FALSE}
USAGE |>   
  group_by(Agency) |>
  summarize(total_vrm = sum(vehicle_revenue_miles, na.rm = T)) |>
  arrange(desc(total_vrm)) |>
  slice(1)
```

Based on this, it can be determined that the [**MTA New York City Transit**](https://new.mta.info/agency/new-york-city-transit) had the most total VRM, of approximately **10,800,000,000 miles**.

Comparatively, a similar code can be used to determine the **transit mode with the most total VRM**: 

```{r echo=TRUE, message=FALSE}
USAGE |>
  group_by(Mode) |>
  summarize(total_vrm = sum(vehicle_revenue_miles, na.rm = T)) |>
  arrange(desc(total_vrm)) |>
  slice(1)
```

It can be determined that [**buses**](https://en.wikipedia.org/wiki/Bus) had the most total VRM, with **49,444,494,088** miles.

Now, a closer look will be had on the MTA, specifically **the number of trips taken on the NYC Subway in May 2024**: 

```{r echo=TRUE, message=FALSE}
USAGE$month <- as.character(USAGE$month)  
USAGE |>
  filter(USAGE$Agency == "MTA New York City Transit",
         USAGE$Mode == "Heavy Rail",
         USAGE$month == "2024-05-01")
```

Here, in May 2024, **180,458,819** trips were taken. 

While this number is indeed high, it would be interesting to analyze the impact the [COVID-19 pandemic](https://www.cdc.gov/covid/about/index.html) had on NYC Subway ridership. To do so, the **fall of NYC subway ridership between April 2019 and April 2020** needs to be determined:

```{r echo=TRUE, message=FALSE}
USAGE |> 
  filter(USAGE$Agency == "MTA New York City Transit", #2019
         USAGE$Mode == "Heavy Rail",
         USAGE$month == "2019-04-01")
USAGE |>
  filter(USAGE$Agency == "MTA New York City Transit", #2020
         USAGE$Mode == "Heavy Rail",
         USAGE$month == "2020-04-01")
```

Based on this information, ridership fell from **232,223,929 in April 2019 to 20,254,269 in April 2020**. This points to a difference of 211,969,660 rides.

## Additional Analysis

Previously, buses were determined to be the transit mode with the most total VRM. It would be interesting to determine the opposite: 

```{r echo=TRUE, message=FALSE}
USAGE |> 
  group_by(Mode) |>   # Look through the lens of transportation mode
  summarize(total_vrm = sum(vehicle_revenue_miles, na.rm = T)) |>   # Obtain the total VRM per mode
  arrange(total_vrm) |>   # Ascending order
  slice(1)   # Obtain the top data point
```

In contrast to buses, [**aerial tramways**](https://en.wikipedia.org/wiki/Aerial_tramway) were the **transit mode with the least total VRM**, with **292,850** miles.

This is a difference of 49,444,201,238 miles. While this might feel drastic, it makes sense. Aerial tramways, vehicles suspended from a system of cables that are propelled through a suspension system, are not commonly used, which attributes to their low mileage. On the other hand, buses are used extensively, with there being numerous routes that run daily, attributing to their high mileage.

Following this contrast, the **agency with the least total VRM** was [**Barnegat Bay Decoy & Baymen's Museum**](https://tuckertonseaport.org/about/), with **2,312** miles.

```{r echo=TRUE, message=FALSE}
USAGE |>
  group_by(Agency) |>   # Look through the lens of transportation agency   
  summarize(total_vrm = sum(vehicle_revenue_miles, na.rm = T)) |>   # Obtain the total VRM per agency
  arrange(total_vrm) |>   # Ascending order
  slice(1)   # Obtain the top data point
```

Compared to the MTA, there is a difference of 10,799,997,688 miles. This large difference can be attributed to the vast difference in sizes between the two. The MTA is the largest transportation system in North America. On the other hand, Barnegat is only a 40 acre cultural center meant to preserve the maritime history of the Jersey Shore. The VRM accumulated correlate to the size and population attributed to each agency.

Because the MTA in NYC has been established as having the largest VRM, examining **which location has the second highest average vehicle revenue miles**, after the New York / New Jersey area, would be interesting. 

```{r echo=TRUE, message=FALSE}
USAGE |> 
  group_by(metro_area) |>   # Look through the lens of location
  summarize(average_vrm = mean(vehicle_revenue_miles, na.rm = T)) |> # Obtain the average VRM per location
  arrange(desc(average_vrm)) |>   # Descending order
  slice(2)   # Obtain the second data point
```

This turns out to be [**Denver, Colorado**](https://www.google.com/maps/place/denver+colorado/data=!4m2!3m1!1s0x876b80aa231f17cf:0x118ef4f8278a36d6?sa=X&ved=1t:242&ictx=111), with **1,565,688** average VRM. This is a 244,360 mile difference to NYC, with 1,810,058 miles.

## Farebox Recovery Analysis

To fully analyze the farebox recovery data, a new table must be created from the previous table, narrowing down the data to specifically 2022. First, the obtained data needs to be edited to create the summary table. 

```{r echo=TRUE, message=FALSE}
## Task 5: Table Summarization
USAGE_2022_ANNUAL <- USAGE |>
  mutate(year = year(month)) |>
  filter(year == 2022) |>
  group_by(NTD_ID, 
           Agency, 
           metro_area, 
           Mode, 
           unlinked_passenger_trips, 
           vehicle_revenue_miles) |>
  summarize(
    total_upt = sum(unlinked_passenger_trips, na.rm = T),
    total_vrm = sum(vehicle_revenue_miles, na.rm = T),
    .groups = "keep",
  ) |>
  ungroup()

USAGE_AND_FINANCIALS <- left_join(USAGE_2022_ANNUAL, 
                                  FINANCIALS, 
                                  join_by(NTD_ID, Mode),
                                  relationship = "many-to-many") |>
  drop_na()
```

Second, the table can be created.

```{r echo=FALSE, message=FALSE}
sample_n(USAGE_AND_FINANCIALS, 1000) |>    # Updated Summary Table
  DT::datatable()
```

\
When analyzing the farebox recovery, it can be determined that the **transit system with the most UPT in 2022** was the **MTA New York CIty Transit**, specifically the **heavy rail**, with 1,793,073,801 trips.

```{r echo=TRUE, message=FALSE}
mostUPT2022 <- USAGE_AND_FINANCIALS |>   # Create a new variable
  group_by(Agency, Mode) |>   # Look through the lens of Agency and Mode
  filter(total_upt > 400000) |>  # Major Transit systems
  summarize(total_upt2022 = sum(total_upt)) |>  # Obtain the total UPT per what was grouped
  arrange(desc(total_upt2022))  # Descending order
head(mostUPT2022, n=1)   # Get only the wanted data
```

However, if examining the **transit system with the highest farebox recovery**, it would be the [**Anaheim Transportation Network**](https://rideart.org/), specifically the **bus**, with 0.865.

```{r echo=TRUE, message=FALSE}
highestfarebox <- USAGE_AND_FINANCIALS |>   # Create a new variable 
  group_by(Agency, Mode) |>   # Look through the lens of Agency and Mode
  filter(total_upt > 400000) |>  # Major Transit systems
  summarize(highestfarebox = sum(`Total Fares`) / sum (Expenses)) |>  # Obtain the ratio of total fares to expenses
  arrange(desc(highestfarebox))  # Descending order
head(highestfarebox, n=1)   # Get only the wanted data
```

Moving on, when examining the **transit system with the lowest expenses per UPT**, it would be the [**University of Georgia**](https://www.uga.edu/) **bus system**, with \$14.90 per trip.

```{r echo=TRUE, message=FALSE}
lowestexpenses <- USAGE_AND_FINANCIALS |>   # Create a new variable
  group_by(Agency, Mode) |>   # Look through the lens of Agency and Mode
  filter(total_upt > 400000) |>  # Major Transit systems
  summarize(lowestexpenses = sum(Expenses) / sum(total_upt)) |>   # Obtain the ratio of expenses to UPT
  arrange(desc(lowestexpenses))   # Descending order
tail(lowestexpenses, n=1)   # Get only the wanted data
```

Comparatively, the **transit system with lowest expenses per VRM** is the [**Interurban Transit Partnership**](https://www.ridetherapid.org/about-the-rapid) **bus system**, with \$84.10 per mile.

```{r echo=TRUE, message=FALSE}
lowestexpensesvrm <- USAGE_AND_FINANCIALS |>   # Create a new variable 
  group_by(Agency, Mode) |>   # Look through the lens of Agency and Mode
  filter(total_upt > 400000) |>  # Major Transit systems
  summarize(lowestexpensesvrm = sum(Expenses) / sum(total_vrm)) |>   # Obtain the ratio of expenses to VRM
  arrange(desc(lowestexpensesvrm))   # Descending order
tail(lowestexpensesvrm, n=1)   # Get only the wanted data
```

Meanwhile, the **transit system with the highest total fares per UPT** is the [**Metro-North Commuter Railroad Company**](https://new.mta.info/agency/metro-north-railroad), with the **bus**, at \$98.70 per trip.

```{r echo=TRUE, message=FALSE}
highestfares <- USAGE_AND_FINANCIALS |>   # Create a new variable 
  group_by(Agency, Mode) |>   # Look through the lens of Agency and Mode
  filter(total_upt > 400000) |>  # Major Transit systems
  summarize(highestfares = sum(`Total Fares`) / sum(total_upt)) |>   # Obtain the ratio of total fares to UPT
  arrange(desc(highestfares))   # Descending order
head(highestfares, n=1)   # Get only the wanted data
```

Comparatively, the **transit system with highest total fares per VRM** is the [**Washington State Ferries**](https://wsdot.wa.gov/travel/washington-state-ferries) with the [**ferryboat**](https://en.wikipedia.org/wiki/Ferry), at \$937 per mile.

```{r echo=TRUE, message=FALSE}
highestfaresvrm <- USAGE_AND_FINANCIALS |>   # Create a new variable
  group_by(Agency, Mode) |>   # Look through the lens of Agency and Mode
  filter(total_upt > 400000) |>   # Major Transit systems
  summarize(highestfaresvrm = sum(`Total Fares`) / sum(total_vrm)) |>  # obtain the ratio of total fares to VRM
  arrange(desc(highestfaresvrm))   # Descending order
head(highestfaresvrm, n=1)   # Get only the wanted data
```

## Conclusions

Overall, the most efficient transit system in the country is the **MTA New York City Transit**. It is the largest North American transit system, and has proven that it is as efficient as its size. It has the **most total and average vehicle revenue miles**. Additionally, in 2022, it was the transit system with the **most unlinked passenger trips**.
